---
title: "Manual Intervention"
author: "André Menezes"
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

# Manual Intervention

## CP6

To illustrate the subjective intervention class we use the CP6 data graphed below. This time series runs from January 1955 to December 1959, providing monthly total sales, in monetary terms on a standard scale, of a product by a major company in UK. Note that the use of standard time series models may not wield satisfactory results as there a some points that need some attention: 

1. During 1955 the market grows fast at a fast but steady rate, 
2. A jump in December 1955. 
3. The sales flattens off for 1956.
4. There is a major jump in the sales level in early 1957.
5. Another jump in early 1958.
6. Throughout the final two years, there is a steady decline back to late 1957.

```{python load-cp6}
cp6 = load_cp6()
```

```{r plot-cp6, echo=FALSE, fig.cap="CP6 sales series", fig.dim=c(16,9), warnings = FALSE, out.width = "90%"}
df_cp6 <- py$cp6 %>% 
  mutate(time = as.Date(time))
p_cp6 <- ggplot(df_cp6, aes(x = time, y = sales)) +
  geom_point(size = 3) +
  scale_y_continuous(breaks = scales::pretty_breaks(10)) +
  scale_x_date(breaks = scales::pretty_breaks(8), date_labels = "%b/%Y") +
  labs(x = "Month/Year", y = "Sales") + 
  theme(axis.text=element_text(size=20), 
  axis.title=element_text(size=20))
p_cp6
```

## Fit Without Intervention

Given this, let's see how a standard dlm performs. The model used is defined below. 

```{python cp6-fit-without-intervention-1}
# Define the growth model
a = np.array([600, 1])
R = np.array([[100, 0], [0, 25]])
mod = dlm(a, R, ntrend=2, deltrend=[0.90, 0.98])
```

```{python cp6-fit-without-intervention-2}
# Filter and Smooth without intervention
smooth = Smoothing(mod=mod)
out_no_int = smooth.fit(y=cp6["sales"])
dict_filter_no_int = out_no_int.get("filter").get("predictive")
```

```{r plot-fit-cp6-filter-no-intervention, echo=FALSE, fig.cap="Mean response for Filtered predictive distribution with $95\\%$ credible interval", fig.dim=c(16,9), warnings = FALSE, out.width = "90%"}
data_predictive = py$dict_filter_no_int 
data_predictive$time = df_cp6$time

data_predictive %>%
  ggplot(aes(x = time, y = y)) +
  geom_point(size = 3) +
  geom_line(aes(y = f), size = 0.8) +
  geom_ribbon(aes(ymin=ci_lower, ymax=ci_upper), 
              alpha=.25, colour = 'grey40') +
  scale_y_continuous(breaks = scales::pretty_breaks(10)) +
  scale_x_date(breaks = scales::pretty_breaks(8), date_labels = "%b/%Y") +
  theme(legend.position = "top") +
  xlab('Month/Year') + ylab('Sales') + 
  theme(axis.text=element_text(size=20), 
  axis.title=element_text(size=20)) +
  colorspace::scale_color_discrete_divergingx()
```

Note that until November 1955 the forecast distribution was quite acceptable, the credible interval was relatively small and the errors was were distributed around zero and inside the interval. But with the jump in December 1955 the level rises dramatically, the biggest problem is not the model's inability to efficiently predict this point, but the influence it has on future predictions. Note that for most of the year 1956 the predicted sales overestimation the real sales, giving a cluster of negative errors $(y_t - f_t)$. In early 1957 another jump was observed, but in this case, it was accompanied by a regime change. And this has great impact in the amplitude of the credible intervals. In early 1958 another regime change, followed by a change in grow, that is not properly modeled since from August 1958 to January 1960 all errors were negative with the exception of July 1959. 

## Fit With Intervention

With the intervention class it is possible to consider outside information to define the prior distribution at the time $t$. This can be done in two ways: noise or subjective. Which must be provided in a list of dictionaries containing the time the intervention will be carried out and the type. Lets start with a empty list

```{python intervention-list-1}
intervention_list = []
```

### Noise Intervention in Prior Variance

In our example, suppose that a change in growth for the year 1956 was anticipated. An increase in uncertainty about level and growth can be done by the addition of a matrix $H_t$ to $R_t$ at time $t=12$ given by

$$
  H_t = \begin{bmatrix}
    100 & 25 \\
    25 & 25
  \end{bmatrix}
$$
Thus, there is an increase (a positive shift) in the prior variance of the components. In our list of interventions we have

```{python intervention-list-2}
intervention_list = [{
  "time_index": 12, "which": ["noise"],
  "parameters": [{
    "h_shift": np.array([0, 0]),
    "H_shift": np.array([[100, 25], [25, 25]])}]
}]
```

where

* `time_index`: time of intervention;
* `which`: type of intervention (in this case, a noise intervention); 
* `parameters`: the values for the intervention.
  + `h_shit`: Shift in mean (we'll see more about that later).
  + `H_shift`: Shift in variance.

### Noise Intervention in Prior Mean and Variance

It is also possible to intervene in the prior mean. Suppose an increase in the market level is expected for the year 1957, we can add a change in level of 80 units and increase the variance by 100 at January $(t=25)$

$$
\mathbf{h}_{25} = \begin{bmatrix}
80 \\ 
0
\end{bmatrix} 
\quad 
\text{and}
\quad 
\mathbf{H}_{25} = \begin{bmatrix}
100 & 0 \\
0 & 0
\end{bmatrix}
$$

now, updating our intervention list 

```{python intervention-list-3}
intervention_list = [{
  "time_index": 12, "which": ["noise"],
  "parameters": [{
    "h_shift": np.array([0, 0]),
    "H_shift": np.array([[100, 25], [25, 25]])}], 
    
  "time_index": 25, "which": ["noise"],
  "parameters": [{
    "h_shift": np.array([80, 0]),
    "H_shift": np.array([[100, 0], [0, 0]])}]
}]
```

In January 1958 $(t=37)$ another jump in level is anticipated, this time of about 100 units with a feeling of increased certainly about the new level and also a anticipated uncertainly for the growth. At this time, the prior mean and variance given by

$$
\mathbf{a}_{37} = \begin{bmatrix}
864.5 \\
0
\end{bmatrix}
\quad 
\text{and}
\quad 
\mathbf{R}_{37} = \begin{bmatrix}
91.7 & 9.2 \\
9.2 & 1.56
\end{bmatrix}
$$

are simply altered to 

$$
\mathbf{a}^{*}_{37} = \begin{bmatrix}
970 \\
0
\end{bmatrix}
\quad 
\text{and}
\quad 
\mathbf{R}^{*}_{37} = \begin{bmatrix}
50 & 0 \\
0 & 5
\end{bmatrix}
$$

### Observational Variance Intervention
It is also possible to perform interventions on observational variance. This can be useful for outlier anticipation. 

Suppose that at the end of 1955 there will be an announcement of future price increases which will result in forward-buying. So, a intervention at December 1955 $(t=12)$ will allow for an anticipated outlier. In late 1956, there is a a view that the marked change in the new year will begin with a maverick value, as the product that are to be discontinued are sold cheaply. 

This interventions can be done by including a variance intervention in our list of interventions for the respective time:

```{python intervention-list-4}
list_interventions = [
    {"time_index": 12, "which": ["variance", "noise"],
     "parameters": [{"v_shift": "ignore"},
                    {"h_shift": np.array([0, 0]),
                     "H_shift": np.array([[1000, 25], [25, 25]])}]
     },
    {"time_index": 25, "which": ["noise", "variance"],
     "parameters": [{"h_shift": np.array([80, 0]),
                     "H_shift": np.array([[100, 0], [0, 0]])},
                    {"v_shift": "ignore"}]},
    {"time_index": 37, "which": ["subjective"],
     "parameters": [{"a_star": np.array([970, 0]),
                     "R_star": np.array([[50, 0], [0, 5]])}]}
]
```

### Performing the fit (filter and smoothing) with interventions
Finally, the fit with intervention can be done using the `Intervention` class. In the `.fit` method we will initialize the model and the loop forecast, observe and update, this time with the interventions given in `list_interventions`, begin. This will return a dictionary with the same structure as presented in the smoothing section.  

```{python cp6-fit-with-intervention}
manual_interventions = Intervention(mod=mod)
out_int = manual_interventions.fit(
    y=cp6["sales"], interventions=list_interventions)
dict_filter_int = out_int.get("filter").get("predictive")
```

```{r plot-fit-cp6-filter-intervention, echo=FALSE, fig.cap="Mean response for filtered predictive distribution with $95\\%$ credible interval and ideal interventions", fig.dim=c(16,9), warnings = FALSE, out.width = "90%"}
data_predictive = py$dict_filter_int 
data_predictive$time = df_cp6$time

data_predictive %>%
  ggplot(aes(x = time, y = y)) +
  geom_point(size = 3) +
  geom_line(aes(y = f), size = 0.8) +
  geom_ribbon(aes(ymin=ci_lower, ymax=ci_upper), 
              alpha=.25, colour = 'grey40') +
  scale_y_continuous(breaks = scales::pretty_breaks(10)) +
  scale_x_date(breaks = scales::pretty_breaks(8), date_labels = "%b/%Y") +
  theme(legend.position = "top") +
  xlab('Month/Year') + ylab('Sales') + 
  theme(axis.text=element_text(size=20), 
  axis.title=element_text(size=20)) +
  colorspace::scale_color_discrete_divergingx()
```
