---
title: "`pybats-detection`: A python package for outlier and structural changes detection in time series analysis"
author: "André Menezes and Eduardo Gabriel"
date: "Last compiled on `r Sys.setlocale(category = 'LC_TIME', locale='en_GB.UTF-8'); format(Sys.time(), '%B %d, %Y')`"
header-includes:
    - \usepackage{setspace}
    - \usepackage{float}
    - \usepackage{geometry}
    - \geometry{a4paper,nohead,left=2.0cm,right=2.0cm,bottom=2.5cm,top=2.5cm}
    - \onehalfspacing
fontsize: 12pt
indent: true
---

```{r setup, include=FALSE}
Sys.setlocale(category = "LC_TIME", locale="en_GB.UTF-8")
options(digits = 4, prompt = ">>> ", continue = ">>> ", width = 70,
        useFancyQuotes = FALSE)
library(reticulate)
use_virtualenv(virtualenv = "pybats_detection",required = TRUE)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "pdf",
  fig.align = "center",
  fig.width = 15,
  fig.height = 6,
  fig.pos = "H",
  prompt = TRUE,
  tidy = TRUE)
```

```{r pkgs, include=FALSE}
library(dplyr)
library(ggplot2)
library(cowplot)
library(knitr)

theme_set(
  theme_cowplot(font_size = 20, font_family = "Palatino") +
    background_grid() +
    theme(text = element_text(size = 20, family = "Palatino"),
          legend.position = "top")
)
blue <- rgb(32, 114, 184, maxColorValue = 255)
red <- rgb(237,0,0, maxColorValue = 255)
```

```{python pkgs-py, include=FALSE}
import numpy as np
import pandas as pd
from scipy import stats
from pybats.dglm import dlm
from pybats_detection.loader import load_cp6
from pybats_detection.loader import load_telephone_calls
from pybats_detection.smooth import Smoothing
from pybats_detection.monitor import Monitoring
from pybats_detection.random_dlm import RandomDLM
from pybats_detection.intervention import Intervention
np.set_printoptions(precision=4)
```


# Introduction

Dynamic Models and Bayesian Forecasting were introduced in the seminal work by
Harrison (1976). Since then, many scientists do seem to have an interest in the
theoretical innovations and practical applications of this technique. 

Two major works in the field of dynamic models are  West and Harrison (1986)
and West and Harrison (1989), in which the authors introduced Bayesian
subjective intervention, automatic monitoring, and adaptation in the class of 
Dynamic Linear Model (DLM).

To our knowledge, there is no open source tool for Bayesian monitoring and
intervention despite the software literature offering a variety of programs to
work with state space models.
To fill this gap, we introduce `pybats-detection` package, an effective
**python** package for the identification of structural changes and outliers
in DLM. The package's current version includes smoothing for univariate DLM,
automatic monitoring, and subjective intervention.


```{r smooth, child = "01__smooth.Rmd"}
```

```{r intervention, child = "02__manual_intervention.Rmd"}
```

```{r monitor, child = "03__automatic_monitor.Rmd"}
```

# References

- Harrison, P.J., Stevens, C.F., 1976. Bayesian forecasting. Journal of the Royal Statistical660
Society. Series B (Methodological) 38, 205--247

- West, M., Harrison, J., 1989. Subjective intervention in formal models. Journal of Forecasting 
8, 33--53. doi:10.1002/for.3980080104.

- West, M., Harrison, P.J., 1986. Monitoring and adaptation in Bayesian forecasting models.
Journal of the American Statistical Association 81, 741--750. doi:10.1080/01621459.1986.10478331.
